---
title: "Formative Assessment"
author: "Dr Irina Chis Ster"
date: "06/03/2021"
output:
  html_notebook:  
    df_print: paged
editor_options:
  markdown:
    wrap: 72
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

# The aim of the formative assessment

The aim of this formative assessment is to prepare you for the exam
associated with this module - Population Health Research.

On this occasion, the data for analyses are similar for all of you.
However, for the formal exam, you will receive different data sets yet
of similar complexity.

This assessment is not formally marked. Nevertheless, if you addressed
it carefully, it would drive you through almost all concepts presented
so far in the lectures as well as through the R procedures which used to
explore them.

As for the meaning of the variables, the information is minimal but
please feel free to look up for more details.

## The data set:

These data are courtesy of Dr John Schorling, Department of Medicine,
University of Virginia School of Medicine. The data are simplified and
consist of 17 variables (some derived for the purpose of this
assessment) on 375 subjects from 1046 subjects who were interviewed in a
study to understand the prevalence of obesity, diabetes, and other
cardiovascular risk factors in central Virginia for African Americans.
The variables in the data are listed below:

-   Id:Subject id

-   Chol: **Cholesterol** is a waxy substance found in the blood. Our
    body needs **cholesterol** to build healthy cells, but high levels
    of **cholesterol** can increase your risk of heart disease or leads
    to the development of fatty deposits in the blood vessels.

-   glu: **Glucose** is the main type of sugar in the blood and is the
    major source of energy for the body's cells. **Glucose** comes from
    the foods we eat or the body can make it from other substances.

-   hdl: High density lipoprotein (HDL) stands for ***high***-***density
    lipoproteins***. It is sometimes called the "good" ***cholesterol***
    because it carries ***cholesterol*** from other parts of your body
    back to your liver. Your liver then removes the ***cholesterol***
    from your body.

-   ratio: Cholesterol/hdl ratio

-   glyhb: Glycosylated hemoglobin (HbA1c) test is an important blood
    test that determine how well you are managing your diabetes.
    Hemoglobin is a substance in red blood cells that carries oxygen to
    tissues. It can also attach to sugar in the blood, forming a
    substance called glycated hemoglobin, or Hemoglobin A1C. The test
    provides an average blood sugar measurement over a 6- to 12-week
    period and is used in conjunction with home glucose monitoring to
    make treatment adjustments.

-   location: A factor with levels Buckingham and Louisa - the
    geographical location of the participants

-   age: Age (years)

-   gender: Gender, male or female

-   bp1s: Systolic blood pressure

-   bp1d: Diastolic blood pressure

-   timeppn: Postprandial time when labs were drawn in minutes = This
    test is done to see how your body responds to sugar and starch after
    you eat a meal. Within two hours of eating, your insulin and blood
    glucose levels should return to normal.

-   timeppn_cat: a variable derived from the above with

    -   1 indicating insulin and blood glucose levels return to normal
        within 1 hour

    -   2 indicating insulin and blood glucose levels return to normal
        between 1 -2 hrs

    -   3 indicating insulin and blood glucose levels return to normal
        between 2 -3 hrs

    -   4 indicating insulin and blood glucose levels return to normal
        between 3 -4 hrs

    -   5 indicating insulin and blood glucose levels return to normal
        after more than 4 hrs

-   bmi: body mass index - continuous variable =the ratio between weight
    (kg) and height^2^ (height in cm)

-   bmi_binary: 1 indicates obese, 0 not obese according to a blood test

-   bmi_cat indicating:

    -   1 normal or underweight (collapsed together as there were very
        few underweight)

    -   2 overweight (\<=25\<bmi\<30)

    -   3 obese (\<=30bmi\<40)

    -   4 excessively obese (bmi\>=40)

-   whr: Waist to hip ratio

There are 3 statistical outcomes of interest - all representing BMI
measurements: **bmi (continuous)**, **bmi_binary** and **bmi_cat
(ordinal/categorical)**. Please consider the rest of the variables as
independent. They may or may not be associated with the different types
of outcomes (bmi). The assignment requires you exactly that: to explore
and interpret the variability in these 3 outcomes using the available
potential explanatory variables and appropriate techniques as taught
throughout the module so far.

## At the start of the assignment:

-   Please run the Miscellaneous R script v6 which should be in the
    Formative assessment folder.
-   Please do not stuck on any technicalities which you may encounter -
    feel free to email me.

## Please read the data in R in the chunk below 

(DiabetesDataModified.txt is a tab separated text file)

```{r}
##reading the data in
mydata <-read.table("DiabetesDataModified.txt", header=T, sep="\t")
names(mydata)
dim(mydata)
```

### 1. Please make a summary of all the variables in the data recognizing their statistical type (1p)

#### Overall (0.33p)

```{r}
summary(mydata) 
```

#### By BMI-binary (0.33p)

```{r}
##ANOVA lecture contains the command!
## summarize the data
by(mydata, mydata$bmi_binary, summary)

```

#### By BMI- as nominal data (summary by each category) (0.33p)

```{r}
##ANOVA lecture contains the command!
## summarize the data
by(mydata, mydata$bmi_cat, summary)

 
```

### 2. Please explore graphically potential relationships between the outcomes and each independent (potential explanatory) variable as you think appropriate. For a full mark you need to produce at least three relevant graphs for each outcome corresponding to a continuous and a categorical/binary independent variable. Please feel free to create more chunks for code (1p).

#### BMI as continuous outcome (0.33p)

```{r}
mydata$location
tabulate(mydata$location)

ggplot(mydata, aes(x=location, y=bmi, fill=location)) + geom_boxplot()+
      scale_color_manual(values=c("#999999", "#E69F00", "#56B4E9"))+
      scale_x_discrete(limits=c("Buckingham", "Louisa"))+
      labs(title = "BMI by location", 
      y = "BMI", x="BMI  ") 
 
```

#### BMI as binary outcome (0.33p)

```{r}
table(mydata$bmi_binary) ##0=male, 1=female
table(mydata$location) ## 0/1= alive/dead


mytable<-table(mydata$bmi_binary, mydata$location)
mytable

margin.table(mytable, 1) # A frequencies (summed over B)
margin.table(mytable, 2) # B frequencies (summed over A)
prop.table(mytable) # cell percentages - joint distribution
prop.table(mytable, 1) # row percentages
prop.table(mytable, 2) # column percentages

tb.prop<-prop.table(mytable, 1) # row percentages
tb.prop
tb.df <- as.data.frame(tb.prop)
tb.df
names(tb.df) <- c("BMIcategory", "Location", "Frequency")
names(tb.df)


ggplot(tb.df, aes(x=BMIcategory, y=Frequency, fill=Location)) + geom_col()
 
```

#### BMI as ordinal outcome (0.33p)

```{r}
table(mydata$bmi_binary) ##0=male, 1=female
table(mydata$location) ## 0/1= alive/dead


mytable<-table(mydata$bmi_cat, mydata$location)
mytable

margin.table(mytable, 1) # A frequencies (summed over B)
margin.table(mytable, 2) # B frequencies (summed over A)
prop.table(mytable) # cell percentages - joint distribution
prop.table(mytable, 1) # row percentages
prop.table(mytable, 2) # column percentages

tb.prop<-prop.table(mytable, 1) # row percentages
tb.prop
tb.df <- as.data.frame(tb.prop)
tb.df
names(tb.df) <- c("BMIcategory", "Location", "Frequency")
names(tb.df)


ggplot(tb.df, aes(x=BMIcategory, y=Frequency, fill=Location)) + geom_col()

ggplot(tb.df, aes(x=BMIcategory, y=Frequency, fill=Location)) + geom_col(position="dodge")
```

### 3. Please investigate the univariate associations between the outcomes and each potential explanatory variable. Please create you own chunks - the advise is to keep them small (3p).

Please indicate the measure of association between the outcome and the
explanatory variable for each outcome type and write down the hypothesis
testing setting for each association (3×0.5p).

Please indicate the p-value associated and the uncertainty of the
estimates which measure each association and **interpret** the findings
(3×0.5p).


### 4. Please use age, gender, cholesterol levels, waist/hip ratio, diastolic and systolic blood pressure to construct a multivariable model for each outcome (5p)

Please conduct model diagnosis assessment and indicate the goodness of
fit for the models in each case (3×0.5p).

Please select the best statistical model using methods presented in the
lectures and comment on your choice (3×0.5p).

Please choose a simple model with two predictors, one continuous
variable and one categorical/binary and produce and visualize
predictions from this model (3×0.5p).

## BMI -continuous outcome

```{r}
mydata$timeppn_cat<-as.factor(mydata$gender)
is.factor(mydata$gender)


reg_model1<-lm(formula = mydata$bmi~mydata$age+mydata$gender+mydata$chol +mydata$whr+mydata$bp1s +mydata$bp1d,
               data = mydata)
summary(reg_model1)


reg_model2<-lm(formula = mydata$bmi~mydata$age+mydata$gender+mydata$chol +mydata$whr+mydata$bp1d  ,
               data = mydata)
summary(reg_model2)


reg_model3<-lm(formula = mydata$bmi~mydata$age+mydata$gender+mydata$whr+mydata$bp1d  ,
               data = mydata)
summary(reg_model3)


reg_model4<-lm(formula = mydata$bmi~ mydata$gender+mydata$whr+mydata$bp1d  ,
               data = mydata)
summary(reg_model4)

lrtest(reg_model1, reg_model2, reg_model3, reg_model4)

confint(reg_model4)

```


```{r}
##checking assumptions -all 4

par(mfrow=c(2,2))
plot(reg_model4)
par(mfrow=c(1,1))

##Cook's distance points

cutoff <-4/(nrow(mydata))
cutoff
plot(reg_model4, which=4, cook.levels=cutoff)
```



```{r}
reg_model4<-lm(formula = mydata$bmi~ mydata$gender+mydata$whr+mydata$bp1d  ,
               data = mydata)
summary(reg_model4)
confint(reg_model4)
```

```{r}
summary(mydata$whr)
summary(mydata$bp1d)

mydata$whr_center<-mydata$whr-0.88
mydata$bp1d_center<-mydata$bp1d -83.38

reg_model5<-lm(formula = mydata$bmi~ mydata$gender+mydata$whr_center+mydata$bp1d_center  ,
               data = mydata)
summary(reg_model5)
confint(reg_model5)

```

Interpretation: The intercept is meaningful and represents the mean BMI  in females (the baseline of the categorical variable) for the mean values of whr and bd1d.
20.03 = The adjusted (for gender and bd1d) increase in BMI for one unit increase in whr 
0.08  = The adjusted (for gender and whr) increase in BMI for one unit increase in bd1d

## BMI binary

```{r}

logitreg_model1<-glm(mydata$bmi_binary~mydata$age+mydata$gender+mydata$chol +mydata$whr+ mydata$bp1s +mydata$bp1d,
               family = "binomial")
summary(logitreg_model1)


logitreg_model2<-glm(formula = mydata$bmi_binary~mydata$age+mydata$gender+mydata$chol +mydata$whr+ mydata$bp1s,
               family = "binomial")
summary(logitreg_model2)


logitreg_model3<-glm(formula = mydata$bmi_binary~mydata$age +mydata$chol +mydata$whr +mydata$bp1s,
              family = "binomial")
summary(logitreg_model3)


logitreg_model4<-glm(formula = mydata$bmi_binary~mydata$age +mydata$chol +mydata$whr ,
              family = "binomial")
summary(logitreg_model4)

lrtest(logitreg_model1,logitreg_model2,logitreg_model3,logitreg_model4)

confint(logitreg_model4)

```



```{r}
summary(mydata$age)
summary(mydata$chol)
summary(mydata$whr)
 

mydata$age_center<-mydata$age -47
mydata$chol_center<-mydata$chol -207.6
mydata$whr_center<-(mydata$whr-0.88)/0.10 ## that is because 1 unit increase in this ratio is a lot

logitreg_model5<-glm(formula = mydata$bmi_binary~mydata$age_center +mydata$chol_center +mydata$whr_center ,
              family = "binomial")
summary(logitreg_model5)

##Hosmer-Lemeshow 
logitgof(mydata$bmi_binary, fitted(logitreg_model5))


exp(coef(logitreg_model5))
exp(cbind(OR = coef(logitreg_model5), confint(logitreg_model5)))

```
Interpretation: 
0.12=the odds of being "obese" for an individual aged 47 years, chol=207 and whr=0.88.

One year increase in age increases the OR of being obese by a factor of 1.04(1.02-1.06) or by 4% - controlling for cholesterol and whr.

One unit increase in cholesterol increases the OR of being obese by a factor of 1.01(1.003-1.016) or by 1% (0.3%-1.6%) - controlling for cholesterol and whr.

```{r}

multinomreg_model1<-multinom(mydata$bmi_cat~mydata$age+mydata$gender+mydata$chol +mydata$whr+mydata$bp1s +mydata$bp1d,
               family = "binomial")
multinomreg_model1_test <- summary(multinomreg_model1)$coefficients/summary(multinomreg_model1)$standard.errors
multinomreg_model1_test
# 2-tailed z test
multinomreg_model1_p <- (1 - pnorm(abs(multinomreg_model1_test), 0, 1)) * 2
multinomreg_model1_p


multinomreg_model2<-multinom(mydata$bmi_cat~mydata$age+mydata$gender+mydata$chol +mydata$whr+mydata$bp1s ,
               family = "binomial")
multinomreg_model2_test <- summary(multinomreg_model2)$coefficients/summary(multinomreg_model2)$standard.errors
multinomreg_model2_test
# 2-tailed z test
multinomreg_model2_p <- (1 - pnorm(abs(multinomreg_model2_test), 0, 1)) * 2
multinomreg_model2_p

lrtest(multinomreg_model2,multinomreg_model1)

multinomreg_model3<-multinom(mydata$bmi_cat~mydata$age+mydata$gender+mydata$chol +mydata$whr ,
               family = "binomial")
multinomreg_model3_test <- summary(multinomreg_model3)$coefficients/summary(multinomreg_model3)$standard.errors
multinomreg_model3_test
# 2-tailed z test
multinomreg_model3_p <- (1 - pnorm(abs(multinomreg_model3_test), 0, 1)) * 2
multinomreg_model3_p


multinomreg_model4<-multinom(mydata$bmi_cat~ mydata$gender+mydata$chol +mydata$whr ,
               family = "binomial")
multinomreg_model4_test <- summary(multinomreg_model4)$coefficients/summary(multinomreg_model4)$standard.errors
multinomreg_model4_test
# 2-tailed z test
multinomreg_model4_p <- (1 - pnorm(abs(multinomreg_model4_test), 0, 1)) * 2
multinomreg_model4_p


multinomreg_model5<-multinom(mydata$bmi_cat~ mydata$gender  +mydata$whr_center ,
               family = "binomial")
multinomreg_model5_test <- summary(multinomreg_model5)$coefficients/summary(multinomreg_model5)$standard.errors
multinomreg_model5_test
# 2-tailed z test
multinomreg_model5_p <- (1 - pnorm(abs(multinomreg_model5_test), 0, 1)) * 2
multinomreg_model5_p


lrtest(multinomreg_model1,multinomreg_model2,multinomreg_model3)
lrtest(multinomreg_model1, multinomreg_model3)
lrtest(multinomreg_model1, multinomreg_model2)
lrtest(multinomreg_model2, multinomreg_model3)

lrtest(multinomreg_model1, multinomreg_model5)
lrtest(multinomreg_model2, multinomreg_model5)
lrtest(multinomreg_model3, multinomreg_model5)
lrtest(multinomreg_model4, multinomreg_model5)

exp(coef(multinomreg_model5))
exp(confint(multinomreg_model5))
 
```
Interpretation:


0.37= This is the relative risk ratio comparing males to females for BMI category 2 relative to BMI baseline given that the other variables in the model are held constant. For  males relative to females, the relative risk for BMI category 2 relative to BMI category 1 would be expected to be 0.37 times smaller given the other variables in the model are held constant. The p-value corresponding to this estimate is 1.025057e-03 and the 95%CI is (0.20, 0.67)

2.10=This is the relative risk ratio for a 0.10 increase in whr for BMI category 2 relative to BMI category 1 given that the other variables in the model are held constant. 

If a subject were to was to have a 0.10 increase in whr, the relative risk for BMI category 2 relative to BMI category 1 would be expected to increase by a factor of 2.10 given the other variables in the model are held constant.
More generally, we can say that if a subject were to increase their whr by 0.10, they would approximately twice more likely expected to fall into BMI category 2 as compared to BMI category 1.


Please comment on the consistency of the findings and indicate which
outcome would you choose in practice and why (1p).

The three most parsimonious (the models with the least number of parameters yet explaining the most variability in the data) models base on the indicated variables pick on different set of predictors.

There might be a couple of reasons for that:
Some important variables might be actually missing from the indicated list (presence of diabetes, hypertension, etc.) 
The thresholds for defining the BMI binary/categorical outcomes might not be adequate.
In general, a continuous outcome is most informative - most detailed. Creating categories based on a continuous variable may be done at the price of loosing/diluting information - hence the message may loose strength.  

